---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Load data
```{r}
tmp<-unzip("activity.zip")
tmp
data<- read.csv(tmp)
str(data)
head(data)
```

Transform data
```{r}
data$date<- as.Date(data$date, "%Y-%m-%d")
str(data)
```


## What is mean total number of steps taken per day?

Calculate the total number of steps taken per day
```{r}
library(plyr)
library(dplyr)
```

```{r}
df1<- ddply(data,.(date),summarize, totalsteps = sum(steps))
df1
```

Make a histogram of the total number of steps taken each day
```{r histogram_of_totalsteps}
par(mfrow=c(1,1))
hist(df1$totalsteps, col= "Green" , xlab = "Number of steps per day", main="Histogram of steps per day")
```

Calculate and report the mean and median of the total number of steps taken per day.
```{r}
mean<- mean(df1$totalsteps, na.rm = TRUE)
mean
median<- median(df1$totalsteps, na.rm = TRUE)
median
```

So, the mean of the total number of steps taken per day is `r mean` and the median is `r median`.


## What is the average daily activity pattern?

Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, 
averaged across all days (y-axis)
```{r average_steps}
averagesteps<- ddply(data,.(interval),summarize, mean = mean(steps, na.rm = TRUE))
head(averagesteps)
with(averagesteps, plot(interval, mean, ylab="Average steps across all days", type="l", main="Average number of
    steps across all days"))
```


Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
averagesteps[averagesteps$mean==max(averagesteps$mean),]
```

So, in the interval 835 or 08:35AM-08:40AM is the maximum number of steps across all days.  
It is logical to think that because in this interval a lot of people go to work,that's why the maximum is in here.

## Imputing missing values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
colSums(is.na(data))
```
We see that only in the column steps there are NA's.
```{r}
totalNA<- sum(is.na(data$steps))
```
The total number of rows with missing values in the dataset is `r totalNA`.


My strategy for filling in all of the missing values in the dataset would be to use the average number 
of steps for each interval to fill in the NAs.

To create a new dataset that is equal to the original dataset but with the missing data filled in,
first I will try to create a new column in the orignal data frame which contains the mean for each interval.
```{r}
df3<- arrange(join(data,averagesteps),interval)
head(df3)
```

Now, I want to sort the data frame by date and then by interval.
```{r}
df3<- arrange(df3, date, interval)
head(df3)
str(df3)
```

I added the column mean to the original data frame.  
My plan is to change the NA values with the mean values for that interval. Because the mean column has numerical values  
and steps column has integer values, I will transform the steps column from an integer vector to a numerical vector.
```{r}
df3$steps<- as.numeric(df3$steps)
str(df3)
```
And now I will change the NA values with the mean values.
```{r}
df3$steps[is.na(df3$steps)]<- df3$mean[is.na(df3$steps)]
head(df3)
```
Finally, I will remove the mean column.
```{r}
newdata<- select(df3, - mean)
head(newdata) 
```

Make a histogram of the total number of steps taken each day and calculate and report the mean   
and median total number of steps taken per day.
```{r histogramNoNA}
df4<- ddply(newdata,.(date),summarize, totalsteps = sum(steps))
df4
hist(df4$totalsteps, col= "Green" , xlab = "Number of steps per day", main="Histogram of steps per day")
```
  
I calculate the new mean and the new median.
```{r}
newmean<- mean(df4$totalsteps)
newmean
newmedian<- median(df4$totalsteps)
newmedian
```
So, the new mean of the total number of steps taken per day is `r newmean` and the new median is `r newmedian`.  
We can see that the new values do not differ from the first ones.  


## Are there differences in activity patterns between weekdays and weekends?

I will create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date 
is a weekday or weekend   
First of all, my system needs the Sys.setlocale command in order to translate days in english.
```{r, results='hide'}
Sys.setlocale("LC_ALL", "English")
```
I will use the ifelse command to create a character vector and then I will transform it to a categorical variable.
```{r}
newdata$day<- ifelse(weekdays(newdata$date) %in% c("Saturday","Sunday"), "weekend", "weekday")
newdata$day<- factor(newdata$day)
head(newdata)
```

I will make a panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, 
 averaged across all weekday days or weekend days.
```{r panelplot}
library(lattice)
dff<-ddply(newdata,.(interval,day),summarize, mean=mean(steps))
head(dff)
xyplot(mean~interval | day, data=dff, ylab="Average number of steps", type="l", layout=c(1,2))
```




